---
title: "Hackathon-IE6600-Sec01-Group5"
author: "Shrunali,Vishnu,Subhasree "
date: "2023-02-12"
output: html_document
editor_options: 
  markdown: 
    wrap: 72
---

<b>Introduction and Problem statement</b>

<p>The music industry has seen significant evolution and growth over the
years, with each decade bringing new styles, artists, and songs to the
forefront. In this data visualization project, we aim to shed light on
the trends and patterns that have emerged in popular music, with a focus
on the different elements that make up a song. Our dataset comprises of
a comprehensive collection of Spotify music data that includes important
attributes such as the song name, artist name, year, popularity, genre,
and various musical features such as danceability, energy, key, and
more.</p>

<p>In the first section of this project, we will analyze the popular
music trends of each decade, exploring the most dominant genres and
artists that have defined the different eras. We also explore
explicitness as a factor in songs over the decades. Our goal is to
provide an overview of the musical landscape of each decade, giving a
better understanding of the cultural and social context in which the
music was produced.</p>

<p>The second section of this project will delve deeper into the
musicality of songs, providing insights into the different elements that
contribute to the overall sound and feel of a song. We will explore
various musical features such as musicality, loudness, speechiness, and
more, to gain a better understanding of how these elements interact and
influence each other.</p>

<p>By utilizing data visualization techniques, we aim to present the
insights and trends of popular music in a visually compelling and
intuitive manner, providing a unique perspective on the history and
evolution of music. This project serves as a valuable resource for music
lovers, researchers, and industry professionals alike, offering a
comprehensive analysis of popular music trends and musical features.</p>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

```{r loading libraries, include=TRUE,message=FALSE,warning=FALSE}
library(knitr)
library(dplyr)
library(tidyr)
library(stringr)
library(tidyverse)
library(reshape2)
library(ggplot2)
library(Hmisc)
library(magrittr)
library(lubridate)
```

```{r reading data, include=TRUE,message=FALSE,warning=FALSE}
df1=read.csv("tracks.csv")
df=read.csv("artists.csv")

```

```{r data cleaning begins, include=TRUE,message=FALSE,warning=FALSE}

#removing the brackets to unnest the the list of values from the artist_id
df1$id_artists = gsub("\\[", "", df1$id_artists)
df1$id_artists = gsub("\\]", "", df1$id_artists)
#splitting using the delimmeter ","
df1$id_artists = strsplit(df1$id_artists, ",")
df3 = tidyr::unnest(df1, id_artists)
df3$id_artists = gsub("'", "", df3$id_artists)

```

```{r using inner join to merge two csv files, include=TRUE,message=FALSE,warning=FALSE}
#using inner join joinning two table, artist and tracks
spotify_data <- inner_join(df3, df, by=c('id_artists'='id' ))

```

```{r  renaming columns in new dataframe, include=TRUE,message=FALSE,warning=FALSE}
spotify_data <-  spotify_data %>% rename("track_id" = "id")
spotify_data <- spotify_data %>% rename("track_name" = "name.x")
spotify_data <- spotify_data %>% rename("track_popularity" = "popularity.x")
spotify_data <- spotify_data %>% rename("track_duration" = "duration_ms")
spotify_data <- spotify_data %>% rename("track_explicit" = "explicit")
spotify_data <- spotify_data %>% rename("track_artist" = "artists")
spotify_data <- spotify_data %>% rename("artist_id" = "id_artists")
spotify_data <- spotify_data %>% rename("track_release_date" = "release_date")
spotify_data <- spotify_data %>% rename("track_danceability" = "danceability")
spotify_data <- spotify_data %>% rename("track_energy" = "energy")
spotify_data <- spotify_data %>% rename("track_key" = "key")
spotify_data <- spotify_data %>% rename("track_loudness" = "loudness")
spotify_data <- spotify_data %>% rename("track_mode" = "mode")
spotify_data <- spotify_data %>% rename("track_speechiness" = "speechiness")
spotify_data <- spotify_data %>% rename("track_acousticness" = "acousticness")
spotify_data <- spotify_data %>% rename("track_instrumentalness" = "instrumentalness")
spotify_data <- spotify_data %>% rename("track_liveness" = "liveness")
spotify_data <- spotify_data %>% rename("track_valence" = "valence")
spotify_data <- spotify_data %>% rename("track_tempo" = "tempo")
spotify_data <- spotify_data %>% rename("track_time_signature" = "time_signature")
spotify_data <- spotify_data %>% rename("artist_followers" = "followers")
spotify_data <- spotify_data %>% rename("artist_genres" = "genres")
spotify_data <- spotify_data %>% rename("artist_name" = "name.y")
spotify_data <- spotify_data %>% rename("artist_popularity" = "popularity.y")
spotify_data_cleaning <- spotify_data
```

```{r data cleaning, include=TRUE,message=FALSE,warning=FALSE}

# adding a new attribute - track duration in minutes 
spotify_data_cleaning <- spotify_data_cleaning %>% mutate(track_time = track_duration/60000)

spotify_data_cleaning <- spotify_data_cleaning %>% rename("track_time_mins" = "track_time")
# checking if any of the rows have all NA values
spotify_data_cleaning[rowSums(is.na(spotify_data_cleaning))!=ncol(spotify_data_cleaning), ]

# data column is not consist and has values like 1922 and 1922-01-01, in order to bring consistency we are adding "-01-01" to the dates given without month and day information
# date data ranges from 1921 - 2020
dates <- spotify_data_cleaning$track_release_date
dates <- as.POSIXct(dates, format = "%Y-%m-%d")

# first convert 1922 to 1922-01-01 and then convert it to date format and specify time zone
date_data <- spotify_data_cleaning$track_release_date
fixed_date <- gsub("1922","1922-01-01",date_data)

spotify_data_cleaning$track_release_date <- ifelse(nchar(spotify_data_cleaning$track_release_date) == 4, paste0(spotify_data_cleaning$track_release_date, "-01-01"), spotify_data_cleaning$track_release_date)

spotify_data_cleaning$track_release_date <- ymd(spotify_data_cleaning$track_release_date)
# release date done cleaning
# adding the changes to the original dataframe
spotify_data <- spotify_data_cleaning

```

```{r scaling track features, include=TRUE,message=FALSE,warning=FALSE}
# since the rest of the attributes have mean != 0, we think the best scaling approach should be min/max scaling
# min_max_scaled <- (data - min(data)) / (max(data) - min(data))
#track_key
key_min <- min(spotify_data_cleaning$track_key)
key_max <- max(spotify_data_cleaning$track_key)
key_range <- key_max - key_min
spotify_data_cleaning$track_key <- (spotify_data_cleaning$track_key - key_min)/key_range

#track_loudness
loudness_min <- min(spotify_data_cleaning$track_loudness)
loudness_max <- max(spotify_data_cleaning$track_loudness)
loudness_range <- loudness_max - loudness_min
spotify_data_cleaning$track_loudness <- (spotify_data_cleaning$track_loudness - loudness_min)/loudness_range

#track tempo
tempo_min <- min(spotify_data_cleaning$track_tempo)
tempo_max <- max(spotify_data_cleaning$track_tempo)
tempo_range <- tempo_max - tempo_min
spotify_data_cleaning$track_tempo <- (spotify_data_cleaning$track_tempo - tempo_min)/tempo_range

# track_time_signature
ts_min <- min(spotify_data_cleaning$track_time_signature)
ts_max <- max(spotify_data_cleaning$track_time_signature)
ts_range <- ts_max - ts_min
spotify_data_cleaning$track_time_signature <- (spotify_data_cleaning$track_time_signature - ts_min)/ts_range

spotify_data_cleaning <- spotify_data_cleaning[complete.cases(spotify_data_cleaning[,"track_release_date"]), ]
# summary(spotify_data_cleaning)
spotify_data <- spotify_data_cleaning
```

##Trend Analysis <b>Question 1: Analyze Spotify's performance since it's
inception in 2006</b>

<p><b>Aim:</b> To provide a visual representation of the distribution of
tracks released by Spotify over different decades from 2006 to 2020.</p>

```{r graph plots begin - donut chart, include=TRUE,message=FALSE,warning=FALSE}
spotify_performance <- spotify_data %>% 
                      mutate(track_release_date = ymd(track_release_date))%>%
                      filter(track_release_date > ymd("2006-01-01"))
spotify_performance[c('Year','Month','Date')]<-str_split_fixed(spotify_performance$track_release_date,"-",3)
spotify_performance_plot <- spotify_performance[,c(2,3)]

spotify_performance <- spotify_performance %>% 
                        group_by( Year) %>%
                        summarise(no_of_songs = n())
spotify_performance <- filter(spotify_performance, Year < "2021")

# Compute percentages
spotify_performance$fraction <- spotify_performance$no_of_songs/sum(spotify_performance$no_of_songs)
# Compute the cumulative percentages (top of each rectangle)
spotify_performance$ymax = cumsum(spotify_performance$fraction)
# Compute the bottom of each rectangle
spotify_performance$ymin = c(0, head(spotify_performance$ymax, n=-1) )
#compute label position
spotify_performance$labelPosition <- (spotify_performance$ymax+spotify_performance$ymin)/2
#compute label
spotify_performance$label <- paste(spotify_performance$Year, "\n value: ", spotify_performance$no_of_songs)
colors=rainbow(15)
ggplot(spotify_performance, aes(ymax = ymax, ymin=ymin, xmax=4, xmin=3,fill=Year))+
  geom_rect()+
  geom_label(x=3.5, aes(y=labelPosition, label=label), size=3.5) +
  coord_polar(theta = "y")+
  xlim(c(2,4))+
  ggtitle("Trends of tracks released on Spotify")+
  theme_void() +
  theme(legend.position = "none") + scale_fill_manual(values = c("#9400D3","#FF69B4","#FFC0CB","#FFA07A","#E0FFFF","#ADD8E6","#87CEFA","#1E90FF","#20B2AA","#90EE90","#00FF7F","#7FFF90","#7FFF00","#DAA520","#FFD700"))
```

<p><b>Conclusion:</b> The aim of this visualization is to provide a
quick and simple way to see how the number of tracks released by Spotify
has changed over the years, broken down by decade. By seeing the size of
each slice, it should be easy to quickly compare the number of tracks
released in different decades and get an overall understanding of the
data. The donut chart depicts the evolution of the number of tracks
released on Spotify, starting with 5304 tracks in 2006 and steadily
rising to 12331 tracks in 2020. However, there was a slight decrease in
the number of tracks released in 2017, which was almost a thousand less
than the previous year.</p>

<b>Question 2:</b>

<p><b>Hypothesis:</b> Covid-19, also known as the novel coronavirus, a
highly infectious respiratory illness swept across the world, affecting
millions of people and causing widespread panic and disruption. The
Covid-19 pandemic has had a profound impact on every aspect of life,
from the economy and public health to the music industry. The pandemic
not only affected physical health but also had a significant impact on
mental health, particularly in terms of depression. Thus, we conducted
an experiment to find the type of tracks released by Spotify in the
years 2019 and 2020.</p>

<p><b>Aim:</b> Do a comparative study for the years 2019 and 2020 in
order to find a relation between the type of songs released on Spotify
and people's mental health.</p>

```{r covid graph boxplot with jitters, include=TRUE,message=FALSE,warning=FALSE}

spotify_pandemic <- spotify_data %>% 
                      mutate(track_release_date = ymd(track_release_date))%>%
                      filter(track_release_date > ymd("2019-01-01") & 
                               track_release_date < ymd("2020-12-31"))
spotify_pandemic <- spotify_pandemic %>% group_by(track_release_date,track_valence)
spotify_pandemic$mood <- ifelse(spotify_pandemic$track_valence > 0.5, "Happy","Sad")
spotify_pandemic$covid_year <- ifelse(spotify_pandemic$track_release_date < "2020-01-01", "Before Covid","Covid")

ggplot(spotify_pandemic, aes(x = covid_year, y = track_valence, colour = mood)) + 
  geom_boxplot(outlier.shape = NA) +
  geom_jitter()+ggtitle("Comparing Emotional Tone in Music Before and During the COVID-19 Pandemic")

```

<p><b>Conclusion</b>: The above graph is a boxplot with jitters plotted
for two time periods - pre-covid and covid. In order to analyze the
relation between type of tracks and the pandemic we have made use of the
metric - valence.</p>

<p>Valence - A measure from 0.0 to 1.0 describing the musical
positiveness conveyed by a track. Tracks with high valence sound more
positive (e.g. happy, cheerful, euphoric), while tracks with low valence
sound more negative (e.g. sad, depressed, angry).</p>

<p>During our analysis we found correlation between valence and other
features. There is significant correlation between the following pairs.
1) valence - danceability 2) valence - loudness 3) valence - energy</p>

<p>Contrary to our belief we see no significant change in the type of
tracks released on Spotify. Hence, our hypothesis was incorrect.</p>

##Feature Analysis

<b>Question 3: How are the features of a song related to one
another?</b>

```{r heatmap - graph-1 relation between song characteristics, include=TRUE,message=FALSE,warning=FALSE,echo=FALSE}

df_sample=spotify_data_cleaning%>%select(track_danceability:track_time_signature)
library(ggplot2)
library(reshape2)
library(plotly)
correlation_matrix <- cor(df_sample)

# Melt the correlation matrix into a dataframe
melted_correlation_matrix <- melt(correlation_matrix)
melted_correlation_matrix<-melted_correlation_matrix%>%rename("Track_characteristics"="Var1")
melted_correlation_matrix<-melted_correlation_matrix%>%rename("Track_characteristics_1"="Var2")
# Plot the heatmap

p<-ggplot(data=melted_correlation_matrix, aes(x=Track_characteristics, y=Track_characteristics_1, fill=value)) + 
  geom_tile(color="black") + geom_text(aes(label = value), color = "black", size = 1)+
  scale_fill_gradientn(colors = hcl.colors(3, "RdYlGn")) + 
  theme_minimal() + 
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, size = 12, hjust=1))+ggtitle("Feature Analysis - Correlation Matrix") 
  coord_fixed()#for square tiles i have use coord_fixes() and heatmap can be formed with the geom_title itself

ggplotly(p)

```

<p>From the above, its observed that track loudness and track energy has
positive correlation.Hence visualized it using scater plot</p>

```{r scatterplot to show correlation between track loudness and energy,include=TRUE,message=FALSE,warning=FALSE}
library(ggplot2)
library(RColorBrewer)
ggplot(data=spotify_data, aes(x=track_loudness, y=track_energy, color=track_mode)) + 
     geom_point(size=1) + 
     # scale_color_gradientn(colors=magma(5)) + 
     theme(legend.position="none")+ggtitle("Correlation between track energy and track loudness")
```

<b>Conclusion</b>

<p>The plot showcases the correlation between different characteristics
of the song tracks, indicating how they are related to one another. If
the values are close to 1, then the features are positively correlated
and are depicted in green, such as between "track_loudness" and
"track_energy." On the other hand, if the values are close to -1, then
the features are negatively correlated and are shown in red, for
instance, between "track_acousticness" and both "track_energy" and
"track_loudness".</p>

<b>Question 4: Now that we know how various features are related or not
related to each other. Understand what drives popularity of a track and
thus from your analysis tell if a new song is released on Spotify will
it be popular or no.</b>

```{r HEXBIN PLOT, include=TRUE,message=FALSE,warning=FALSE}
spotify_track_artist <- spotify_data[,c(3,21,24)]
# colnames(spotify_track_artist)
# Load the hexbin library 
library(hexbin)
# Create the hexbin object
hexbin_obj <- hexbin(y = spotify_track_artist$track_popularity, x = spotify_track_artist$artist_popularity, xbins = 15)

# Create the hexbin plot

ggplot(data = spotify_track_artist, aes(x=spotify_track_artist$artist_popularity,y=spotify_track_artist$track_popularity))+geom_hex(bins=18)+
  scale_fill_gradient(low = "white", high = "red")+
  labs(title = " Hexbin Plot for track popularity vs artist popularity", x="artist popularity",y = "track popularity")
```

<b>Conclusion:</b>

<p>From the heatmap used to analyze different track features we came to
a conclusion that the track popularity is highly dependent on the
popularity of an artist and how many followers he/she has. This Hex Bin
plot helps us in understanding this relation further.</p>

<p>Hexbin plots take in lists of X and Y values and returns what looks
somewhat similar to a scatter plot, the entire graphing space has been
divided into hexagons (like a honeycomb) and all points have been
grouped into their respective hexagonal regions with a color gradient
indicating the density of each hexagonal area.</p>

<p>From this graph we undestand that highest majority of the artists
have very low or negligible popularity and hence popularity of their
tracks is low. At the same time, artists with high popularity have hit
tracks. Thus if a new song is released on Spotify, given that the track
artist is popular the track has a good chance of becoming a hit.</p>

<b>Question 5: What factors contribute to the tracks popularity of an
popular artist?</b>

```{r , include=TRUE,echo=FALSE,message=FALSE,warning=FALSE}

df_s=spotify_data_cleaning%>%filter(artist_popularity==max(artist_popularity))
df_s[c('Year','Month','Date')]<-str_split_fixed(df_s$track_release_date,"-",3)
df_s

df_s$track_popularity <- df_s$track_popularity /100
p1<-ggplot(df_s, aes(x =track_popularity , y = track_loudness)) +
  geom_line() +
  scale_color_discrete(name = "track_popularity") +
  xlab("track_popularity") +
  ylab("Variable 2") +
  ggtitle("Scatter Plot of track_popularity and track_loudness")+ geom_smooth(formula = y ~ x,method = "lm")
p2<-ggplot(df_s, aes(x =track_popularity , y = track_danceability)) +
  geom_line() +
  scale_color_discrete(name = "track_popularity") +
  xlab("track_popularity") +
  ylab("track_danceability") +
  ggtitle("Scatter Plot of track_popularity and track_danceability")+ geom_smooth(formula = y ~ x,method = "lm")
p3<-ggplot(df_s, aes(x =track_popularity , y = track_energy)) +
  geom_line() +
  scale_color_discrete(name = "track_popularity") +
  xlab("track_popularity") +
  ylab("track_energy") +
  ggtitle("Scatter Plot of track_popularity and track_energy")+ geom_smooth(formula = y ~ x,method = "lm")
p4<-ggplot(df_s, aes(x =track_popularity , y = track_valence)) +
  geom_line() +
  scale_color_discrete(name = "track_popularity") +
  xlab("track_popularity") +
  ylab("track_valence") +
  ggtitle("Scatter Plot of track_popularity and track_valence")+ geom_smooth(formula = y ~ x,method = "lm")
p5<-ggplot(df_s, aes(x =track_popularity , y = track_tempo)) +
  geom_line() +
  scale_color_discrete(name = "track_popularity") +
  xlab("track_popularity") +
  ylab("track_tempo") +
  ggtitle("Scatter Plot of track_popularity and track_tempo")+ geom_smooth(formula = y ~ x,method = "lm")
p6<-ggplot(df_s, aes(x =track_popularity , y = track_instrumentalness)) +
  geom_line() +
  scale_color_discrete(name = "track_popularity") +
  xlab("track_popularity") +
  ylab("track_instrumentalness") +
  ggtitle("Scatter Plot of track_popularity and track_instrumentalness")+ geom_smooth(formula = y ~ x,method = "lm")
p7<-ggplot(df_s, aes(x =track_popularity , y = track_acousticness)) +
  geom_line() +
  scale_color_discrete(name = "track_popularity") +
  xlab("track_popularity") +
  ylab("track_acousticness") +
  ggtitle("Scatter Plot oftrack_popularity and track_acousticness")+ geom_smooth(formula = y ~ x,method = "lm")
p8<-ggplot(df_s, aes(x =track_popularity , y = track_speechiness)) +
  geom_line() +
  scale_color_discrete(name = "track_popularity") +
  xlab("track_popularity") +
  ylab("track_speechiness") +
  ggtitle("Scatter Plot of Two Continuous Variables and track_speechiness")+ geom_smooth(formula = y ~ x,method = "lm")
p9<-ggplot(df_s, aes(x =track_popularity , y = track_mode)) +
  geom_line() +
  scale_color_discrete(name = "track_popularity") +
  xlab("track_popularity") +
  ylab("track_mode") +
  ggtitle("Scatter Plot of track_popularity and track_mode")+ geom_smooth(formula = y ~ x,method = "lm")
p10<-ggplot(df_s, aes(x =track_popularity , y = track_key)) +
  geom_line() +
  scale_color_discrete(name = "track_popularity") +
  xlab("track_popularity") +
  ylab("track_key") +
  ggtitle("Scatter Plot of track_popularity and track_key")+ geom_smooth(formula = y ~ x,method = "lm")
p11<-ggplot(df_s, aes(x =track_popularity , y = track_liveness)) +
  geom_line() +
  scale_color_discrete(name = "track_popularity") +
  xlab("track_popularity") +
  ylab("track_liveness") +
  ggtitle("Scatter Plot of track_popularity and track_liveness")+ geom_smooth(formula = y ~ x,method = "lm")
library(gridExtra)
library(grid)
grid.arrange(p1,p2, p3, p4 ,p5,p6,p7,p8,p9,p10,p11,nrow=4, ncol=3)
```

<p>The line graph indicates that track characteristics do not play a
significant role in determining the popularity of an artist's tracks</p>

<b>

<p>Question 6: What is the occurrence rate of track names for a leading
artist?</p>

</b>

```{r graph-4 analysing frequency of tracks names of an popular artist over the years using wordcloud,include=TRUE,message=FALSE,warning=FALSE}
df_s_1=spotify_data_cleaning
df_s_1[c('Year','Month','Date')]<-str_split_fixed(df_s_1$track_release_date,"-",3)
df_s_1<-df_s_1%>%filter(artist_popularity==max(artist_popularity))
df_s_1_grp<-df_s_1%>%filter(artist_popularity==max(artist_popularity))%>%group_by(track_name)%>%summarise(total_popularity=sum(track_popularity))
c<-df_s_1%>%group_by(track_name)%>%summarise(num=n())
df_s_1 <- df_s_1 %>%
    mutate(track_name = gsub(" ", "_", track_name))

#install.packages("tm")
library(tm)
#Create a vector containing only the text
text <- df_s_1$track_name
# Create a corpus  
docs <- Corpus(VectorSource(text))
dtm <- TermDocumentMatrix(docs) 
matrix <- as.matrix(dtm) 
words <- sort(rowSums(matrix),decreasing=TRUE) 
df <- data.frame(word = names(words),freq=words)

set.seed(1234) # for reproducibility 

#install.packages("wordcloud")
library(wordcloud)
#install.packages("RColorBrewer")
library(RColorBrewer)

suppressWarnings(wordcloud(words = df$word, freq = df$freq, min.freq = 1,
                           max.words=500, random.order=FALSE, rot.per=0.35,
                           colors=brewer.pal(8, "Dark2")))
```

<b>Conclusion</b>

<p>The wordcloud visualization indicates that the track titles "hold_on"
and "anyone" appear frequently for the artist Justin Bieber. The font
size in the wordcloud allows us to determine which track titles occur
most frequently and which do not. Hence, the wordcloud visual provides
information about the occurrence rate of track names for a leading
artist.</p>

<b>

<p>Question 7: What variation in track characteristics can be observed
across the top 10 music genres?</p>

</b>

```{r graph-2 stacked bar graph how the track characteristics changes based on genres(top),include=TRUE,message=FALSE,warning=FALSE}

library(dplyr)
library(tidyr)
library(stringr)
library(tidyverse)
library(reshape2)

genre_data <- spotify_data_cleaning%>%filter(artist_genres!="[]")%>%group_by(artist_genres)%>%summarise(sum_popularity = sum(track_popularity))%>%top_n(10, sum_popularity)%>%arrange(desc(sum_popularity))%>%rename(genres=artist_genres)

df_sample=spotify_data_cleaning%>%select('track_danceability','track_energy','track_key','track_loudness','track_mode','track_speechiness','track_acousticness','track_instrumentalness','track_liveness','track_valence','track_tempo','track_time_signature','artist_genres')
long_df <- melt(df_sample, id.vars = "artist_genres")
grouped_df <- long_df %>%group_by(artist_genres,variable) %>%summarise(average = mean(value))

df_sample_1<-inner_join( genre_data,grouped_df, by=c('genres'='artist_genres' ))

colors <- c("#0429b3", "#2853b2", "#3c76b1", "#5199b0", "#65bcaf", "#79dfae", "#8df2ad", "#a1f5ac", "#b5f8ab", "#c9fbaa", "#deffa9", "#f2ffa8")

#colors <- c("#0a3d91", "#186faf", "#2da4d4", "#48c7e0", "#63e1eb", "#7ff5f5", "#9bf9ff", "#b7feff", "#d2ffff", "#edffff", "#f7f7f7", "#ffffff")

p<-ggplot(df_sample_1, aes(x =genres, y = average, fill = variable)) +
  geom_bar(stat = "identity") +
  ggtitle("Stacked Bar Plot of Two Categorical Variables and One Continuous Variable") +
  xlab("genres") +
  ylab("average")+
  theme(axis.text.x = element_text(angle = 90, vjust = 1, hjust=1))+scale_fill_manual(values = colors)
ggsave("plot.png", p, width = 10, height = 10, units = "in")
p

```

<b>Conclusion</b>

<p>According to the Stacked bar chart, it can be noted that the "kleine
hoerspie" genre possesses the highest values for several track features,
including "track_danceability," "track_energy," "track_key,"
"track_loudness," "track_mode," "track_speechiness," and
"track_acousticness." This allows us to observe the variations in each
track characteristic among the top 10 music genres.</p>

##Time Series Analysis <b>Question 8: What is the inclination of Spotify
users towards particular artists, and does the year of release play a
role in shaping these preferences?</b>

```{r,include=TRUE,message=FALSE,warning=FALSE}

spotify_data$track_release_year = substr(spotify_data$track_release_date, 1, 4)

spotify_data$track_release_year<- as.numeric(spotify_data$track_release_year)

library(ggplot2)
library(dplyr)

# Group the data by decades
spotify_data_grouped <- spotify_data %>%
  mutate(track_release_decade = as.integer(floor(track_release_year / 10) * 10)) %>%
  group_by(track_release_decade)

spotify_data_grouped$track_release_decade<- as.numeric(spotify_data_grouped$track_release_decade)
spotify_data_grouped<- spotify_data_grouped[complete.cases(spotify_data_grouped), ]
spotify_data_grouped <- subset(spotify_data_grouped,track_release_decade!=1900 )
```

```{r,include=TRUE,message=FALSE,warning=FALSE}
library(ggplot2)
df_filtered_1 <- spotify_data_grouped %>% 
  distinct(artist_name, .keep_all = TRUE)
# Get the top 5 artists by popularity for each decade
df_top5_artist <- df_filtered_1 %>%
  group_by(track_release_decade) %>%
  top_n(5, artist_popularity) %>%
  ungroup()
# Plot the dot plot
ggplot(df_top5_artist, aes(x = reorder(artist_name, track_release_decade), y = artist_popularity)) +
  geom_point(aes(color = as.factor(track_release_decade))) +
  scale_color_discrete(name = "Release Decade") +
  ggtitle("Top 5 Unique Artists by Popularity for Each Decade (1920-2020)") +
  xlab("Artist Name") +
  ylab("Popularity") +
  coord_flip()
```

<b>Conclusion</b>

<p>We plot the top 5 artists in each decade between 1920-2020. This plot
also shows an increase in popularity for artists in more recent years
compared to those in the early ones.This could be due to a few different
factors such as the evolution of the music industry in terms of genre
preferences, better technology used to create music, globalization of
popular genres, etc.</p>

<b> Question 9: What is the inclination of Spotify users towards
different genres of music? Does the decade that this genre originated
have an impact on whether people still listen to it?</b>

<p>

```{r,include=TRUE,message=FALSE,warning=FALSE}
genre_data <- spotify_data_grouped %>%
  mutate(artist_genres = str_remove_all(artist_genres, "[\\[\\]']")) %>%
  separate_rows(artist_genres, sep = ",")

genre_data$artist_genres <- ifelse(genre_data$artist_genres == "", NA, genre_data$artist_genres)

genre_data<- genre_data[complete.cases(genre_data), ]

genre_data <- group_by(genre_data, track_release_decade, artist_genres)
genre_data <- summarise(genre_data, popularity = mean(artist_popularity))

genre_data <- genre_data %>%
  group_by(track_release_decade) %>%
  top_n(5, popularity)
```

```{r,include=TRUE,message=FALSE,warning=FALSE}

ggplot(genre_data, aes(x = track_release_decade, y = reorder(artist_genres, track_release_decade), fill = popularity)) + 
  geom_tile() + 
  xlab(label ="Decade")+
  ylab(label ="Genre")+
  scale_x_continuous(breaks=seq(1920, 2020, by=10)) +
  scale_fill_gradient(name = "Popularity",low = "green", high = "black") + 
  theme_minimal()

```

<b>Conclusion</b>

<p>We plot the top 5 genres by popularity that originated in each decade
between 1920-2020. What is interesting to note is that the most popular
genres of all time on Spotify are those that originated in the 2000s.
From the previous plot(PLOT1), we saw that artists of older decades did
not fare well in terms of popularity, but the same trend does not apply
to genres. This shows that artists of recent years have adapted older
genres into their style of music to make it more appealing to the
audience.</p>

<b> Question 10: What is the popular attitude towards explicit content
in music?</b>

```{r,include=TRUE,message=FALSE,warning=FALSE}
grouped_data <- spotify_data_grouped %>% 
  group_by(track_release_decade) %>% 
  summarise(mean_popularity = mean(track_popularity),
            mean_explicit = mean(track_explicit))

ggplot(grouped_data, aes(x = track_release_decade, y = mean_explicit, fill = mean_popularity)) + 
  geom_col(show.legend = FALSE) + 
  scale_fill_gradient(low = "green", high = "black") +
  scale_x_continuous(breaks=seq(1920, 2020, by=10)) +
  labs(x = "Decade", y = "Mean Explicitness", 
       title = "Explicitness of Music Over the Decades",
       subtitle = "1920-2020") + 
  theme_minimal()
```

<p>

To understand this better, we plot the mean for explicitness over the
decades from 1920-2020. The mean in this case would represent the
proportion of 1s (explicit) in the data set. So, for a given decade, if
the mean of the explicitness variable is 0.6, this would mean that 60%
of the tracks in the data set for that decade are explicit. By
aggregating the mean explicitness per decade, we can see how the
proportion of explicit tracks has changed over time and whether there
are any trends or shifts in attitudes towards explicit content.This
shows us that the cultural attitude towards explicit music and an
artist's freedom of expression has increased largely, with most popular
songs in the recent decades of the 2000s having popular tracks with
explicit lyrics.

<p>

<p>To further understand the portion of explicit lyrics in top songs of
the artists that have dominated each decade of release, we plot the
explicitness for the top song of each artist by decade.</p>

```{r,include=TRUE,message=FALSE,warning=FALSE}
top5_artist_song <- df_top5_artist %>% 
  group_by(track_release_decade, artist_name) %>% 
  top_n(1,track_popularity) %>% 
  ungroup()
top5_artist_song$track_explicit <- ifelse(top5_artist_song$track_explicit == 0, "no", "yes")


ggplot(top5_artist_song, aes(x = track_release_decade, y = reorder(artist_name, track_release_decade), color = track_explicit)) + 
  geom_point(size = 2) + 
  scale_color_manual(values = c("green", "black"), guide = FALSE) + 
  labs(x = "Decade", y = "Artist Name", color = "Explicitness") + 
  ggtitle("Explicitness of the Top Songs of Top 5 Artists based on the Decade the song was released") + 
  theme_minimal()


```

<b>Conclusion </b>

<p>We notice that most popular artists of recent decades do not in fact
have explicit lyrics in their top tracks. This shows us that though
explicit music is popular, it is not a necessity for the success of the
song. This could be due to various factors such as the demographic of
the audience, Spotify's recommendation algorithms, etc.</p>

<b>Report Conclusion: </b>

<p>

In conclusion, our data visualization project based on Spotify's track
data has revealed several interesting insights into the music industry.
Through the use of various graphical representations, we were able to
understand the distribution of different audio features of songs, the
popularity of different genres, and the trend of listening behavior over
time. Our findings showed that danceability, energy, and tempo are
important factors that contribute to the popularity of a song, while
hip-hop and pop music are the most streamed genres on Spotify. Moreover,
we observed a steady growth in the number of streams over the years,
indicating an increasing trend in the usage of Spotify and digital music
streaming services.

</P>

<p>In summary, this project has provided a comprehensive view of the
music industry and the trends that are shaping it. By utilizing data
visualization techniques, we have been able to effectively communicate
complex information and derive meaningful insights. We hope that these
findings will be useful for artists, music labels, and other
stakeholders in the industry to make informed decisions and shape the
future of music.</p>
